Phylogenetics and the human microbiome.

### 2013-09-30 sbreview.tex
Edits from excellent feedback by @nhoffman.

\subsection{Microbial community estimation using marker gene surveys}
Our modern knowledge of the microbial world is due to a large part derived from the methods of Carl Woese and colleagues who pioneered the use of marker genes as a way to distinguish between microbial lineages \citep{fox1977comparative}.
Their work, and the scientists who followed them, focused on the 16S ribosomal gene as a genetic marker.
This gene was chosen because it has regions of high and low diversity, which enable resolution on a variety of evolutionary time scales.
Regions of low diversity in 16S also enabled the development of the first ``universal'' 16S PCR primers \citep{lane1985rapid} which enabled surveys of all organisms regardless of whether they can be cultured.

@@

Despite the high throughput and low cost of modern sequencing, inherent challenges remain for the use of population census by marker gene sequencing.
Most fundamentally, various microbes have different DNA extraction efficiencies, even with stringent protocols, meaning that the representation of marker gene sequences is not representative of the actual communities \citep{morgan2010metagenomic}.
Current sequencing technology is limited to a length that is shorter than most genes, which limits the resolution of the analyses.
``Primer bias,'' or differing amplification levels of various sequences based on their affinity for the primers \citep{suzuki1996bias,polz1998bias}, is a challenge and has led to the standardization of primers \citep{methe2012framework}.
Worse, multiplex PCR is known to create chimeric (i.e.\ spurious recombinant) sequences via partial PCR products \citep{hugenholtz2003chimeric,ashelford2005least,haas2011chimeric,schloss2011reducing}.
Correspondingly, chimera checking software has been developed \citep[including][]{ashelford2006new,edgar2011uchime}.
Also, 16S can be present in up to 15 copies and there can be diversity within the copies \citep{klappenbach2001rrndb}.
Recent work by \citet{kembel2012incorporating} implements the independent contrasts \citep{felsenstein1985phylogenies} method to correct for copy number, which has been helpful despite a moderate evolutionary signal in copy number variation \citep{klappenbach2000rrna}.
Some groups have reported advantages to using alternate single-copy genes as markers for characterization of microbial communities \citep[e.g.][]{case2007rpob,mcnabb2004hsp65}, however 16S remains the dominant locus used by a large margin.

\subsection{Metagenomes}
As described above ``metagenome" means that data is sheared randomly across the genome rather than amplified from a specific region, and thus the genetic region of a read is unknown in addition to the organism it was derived from.
Because it does proceed through an amplification step, it does not have the same PCR primer biases as a marker gene survey, although sequencing is known to have biases.

@@

An obvious application of phylogenetics is to perform taxonomic classification, as the taxonomy is at least in part defined by phylogeny.
However, comparisons of taxonomic classification programs \citep{liu2008accurate,bazinet2012comparative} have indicated that current implementations of phylogenetic methods do not perform as well as simple classifiers based on $k$-mer composition \citep{wang2007naive,rosen2008metagenome}.
% \citep{werner2011impact} Impact of training sets on classification of high-throughput bacterial 16s rRNA gene surveys
Some authors report that a combination of composition-based and homology-based classifiers work best \citep{brady2009phymm,parks2011classifying}.
The MEGAN program \citep{huson2007megan,huson2011integrative} uses a least common ancestor (LCA) strategy on taxonomies to derive taxonomic assignments for an unknown sequence from BLAST results of an unknown sequence onto a database of sequences with those taxonomic labels.
\citet{munch2008statistical,munch2008fast} infer taxonomic assignment by automatically retrieving sequences equipped with taxonomic information and building a tree on them along with an unknown sequence.
\citet{segata2012metagenomic} propose a clever approach to inferring organisms present in a metagenomic sample from a metagenome by compiling a database of clade-specific genes, then classifying a given read as being from the only clade that has the corresponding gene.
They show that this has good sensitivity and specificity, however, this method can only be used to classify to organisms whose genome has been sequenced.

\subsection{The role of Operational Taxonomic Units (OTUs)}
Although there continues to be a lively debate on if there is a meaningful concept of species for microbes \citep{bapteste2009prokaryotic,caro2012bacterial}, a substantial part of human microbiome research has replaced any traditional species concept with the notion of Operational Taxonomic Units (OTUs).
An OTU is a proxy species concept that is typically defined with a fixed divergence cutoff, most commonly at 97\% sequence identity, such that each OTU consists of sequences that are closer to each other than that cutoff.
It is common for trees to be built on sequence representatives from these OTUs, and the abundance of an OTU to be given by the number of sequences that sit within that cluster.
I briefly describe the mini-industry of OTU clustering techniques to contrast with the phylogenetic literature on species delimitation \citep{pons2006sequence,yang2010bayesian}.

@@

There are many OTU clustering methods of various speeds and strategies.
Of the fixed-cutoff methods, the traditional choice has been CD-HIT \citep{li2006cdhit}, which seems to have been supplanted by clustering features of USEARCH \citep{edgar2010usearch} and perhaps now its descendant UPARSE \citep{edgar2013uparse}.
These methods are very fast and are heuristic in that they are described as an algorithm rather than as global optimization of some notion of goodness of clustering.
Different ways of doing this heuristic clustering can result in very different results \citep{white2010alignment}.
There are also methods that try to come up with sequence groupings that more closely reflect the type of species divisions found in taxonomies \citep{navlakha2009finding}.
\citet{hao2011clustering} use a Gaussian mixture model formulation for clustering to avoid fixed cutoff values.

The centrality of the OTU concept can be seen by the fact that the contingency table of OTU observations is considered to be the fundamental data type for 16S studies \citep{mcdonald2012biological}, or that methods have been devised to find OTUs from non-overlapping sequences \citep{sharpton2011phylotu}.
With the exception of this last reference, OTU inference is not considered to be a phylogenetic problem but rather something to be performed before phylogenetic inference begins.

@@

@@

Opportunities for clinical applications may present themselves, but only some of these will be sequencing based.
For example, routine 16S sequencing is likely soon to be completely replaced by Matrix-Assisted Laser Desorption Ionization--Time of Flight (MALDI-TOF) mass spectrometry for identification of a single known microbe grown in culture \citep{clark2013matrix}.
However, for diagnoses that are on the level of microbial communities, sequencing and consequent analysis methods will still be required \citep[see review in][]{Rogers2013271}.
Inexpensive whole-genome sequencing will certainly have a profound impact on clinical practice and epidemiological studies \citep{didelot2012transforming}.

>>>>

\subsection{Microbial community estimation using marker gene surveys}
Our modern knowledge of the microbial world is due to a large part derived from the methods of Carl Woese and colleagues who pioneered the use of marker genes as a way to distinguish between microbial lineages \citep{fox1977comparative}.
Their work, and the scientists who followed them, focused on the 16S ribosomal gene (henceforth simply ``16s'') as a genetic marker.
This gene was chosen because it has regions of high and low diversity, which enable resolution on a variety of evolutionary time scales.
Regions of low diversity in 16S also enabled the development of the first ``universal'' 16S PCR primers \citep{lane1985rapid} which enabled surveys of almost all bacteria regardless of whether they can be cultured.

@@

Despite the high throughput and low cost of modern sequencing, inherent challenges remain for the use of population census by marker gene sequencing.
Most fundamentally, various microbes have different DNA extraction efficiencies, even with stringent protocols, meaning that the representation of marker gene sequences is not representative of the actual communities \citep{morgan2010metagenomic}.
Current sequencing technology is limited to a length that is shorter than most genes, which limits the resolution of the analyses.
``Primer bias,'' or differing amplification levels of various sequences based on their affinity for the primers \citep{suzuki1996bias,polz1998bias}, is a challenge and has led to the standardization of primers \citep{methe2012framework}.
Worse, multiplex PCR is known to create chimeric (i.e.\ spurious recombinant) sequences via partial PCR products \citep{hugenholtz2003chimeric,ashelford2005least,haas2011chimeric,schloss2011reducing}.
Correspondingly, chimera checking software has been developed \citep[including][]{ashelford2006new,edgar2011uchime}.
Also, 16S can be present in up to 15 copies and there can be diversity within the copies \citep{klappenbach2001rrndb}.
Recent work by \citet{kembel2012incorporating} implements the independent contrasts \citep{felsenstein1985phylogenies} method to correct for copy number, which has been helpful despite a moderate evolutionary signal in copy number variation \citep{klappenbach2000rrna}.
Some groups have reported advantages to using alternate single-copy genes as markers for characterization of microbial communities \citep[e.g.][]{case2007rpob,mcnabb2004hsp65}, however 16S remains the dominant locus used by a large margin.
A final cause of noise is next-generation sequencing error: this is certainly a problem for surveys and metagenomes, but is becoming less of a problem as technology improves.
I will not address it specifically except for in the inference of operational taxonomic units as described below.

\subsection{Metagenomes}
As described above ``metagenome" means that data is sheared randomly across the genome rather than amplified from a specific region, and thus the genetic region of a read is unknown in addition to the organism it was derived from.
Because it does not proceed through an amplification step, it does not have the same PCR primer biases as a marker gene survey, although sequencing is known to have biases.

@@

An obvious application of phylogenetics is to perform taxonomic classification, as the taxonomy is at least in part defined by phylogeny.
However, comparisons of taxonomic classification programs \citep{liu2008accurate,bazinet2012comparative} have indicated that current implementations of phylogenetic methods do not perform as well as simple classifiers based on $k$-mer composition \citep{wang2007naive,rosen2008metagenome}.
\citet{srinivasan2012bacterial} find that phylogenetic methods to do taxonomic classification can outperform composition-based techniques at least for certain taxonomic groups.
% \citep{werner2011impact} Impact of training sets on classification of high-throughput bacterial 16s rRNA gene surveys
Some authors report that a combination of composition-based and homology-based classifiers work best \citep{brady2009phymm,parks2011classifying}.
The MEGAN program \citep{huson2007megan,huson2011integrative} uses a least common ancestor (LCA) strategy on taxonomies to derive taxonomic assignments for an unknown sequence from BLAST results of an unknown sequence onto a database of sequences with those taxonomic labels.
\citet{munch2008statistical,munch2008fast} infer taxonomic assignment by automatically retrieving sequences equipped with taxonomic information and building a tree on them along with an unknown sequence.
\citet{segata2012metagenomic} propose a clever approach to inferring organisms present in a metagenomic sample from a metagenome by compiling a database of clade-specific genes, then classifying a given read as being from the only clade that has the corresponding gene.
They show that this has good sensitivity and specificity, however, this method can only be used to classify to organisms whose genome has been sequenced.

\subsection{The role of Operational Taxonomic Units (OTUs)}
Although there continues to be a lively debate on if there is a meaningful concept of species for microbes \citep{bapteste2009prokaryotic,caro2012bacterial}, a substantial part of human microbiome research has replaced any traditional species concept with the notion of Operational Taxonomic Units (OTUs).
An OTU is a proxy species concept that is typically defined with a fixed divergence cutoff, most commonly at 97\% sequence identity, such that each OTU consists of sequences that are closer to each other than that cutoff.
It is common for trees to be built on sequence representatives from these OTUs, and the abundance of an OTU to be given by the number of sequences that sit within that cluster.
I briefly describe the mini-industry of OTU clustering techniques to contrast with the phylogenetic literature on species delimitation \citep{pons2006sequence,yang2010bayesian}.
I will use the term \emph{OTU inference} despite the fact that there is no clear definition of the OTU concept.

@@

There are many OTU inference methods of various speeds and strategies.
Of the fixed-cutoff methods, the traditional choice has been CD-HIT \citep{li2006cdhit}, which seems to have been supplanted by clustering features of USEARCH \citep{edgar2010usearch} and perhaps now its descendant UPARSE \citep{edgar2013uparse}.
These methods are very fast and are heuristic in that they are described as an algorithm rather than as global optimization of some notion of goodness of clustering.
Different ways of doing this heuristic clustering can result in very different results \citep{white2010alignment}.
There are also methods that try to come up with sequence groupings that more closely reflect the type of species divisions found in taxonomies \citep{navlakha2009finding}.
\citet{hao2011clustering} use a Gaussian mixture model formulation for clustering to avoid fixed cutoff values.

The centrality of the OTU concept can be seen by the fact that the contingency table of OTU observations is considered to be the fundamental data type for 16S studies \citep{mcdonald2012biological}, or that methods have been devised to find OTUs from non-overlapping sequences \citep{sharpton2011phylotu}.
A significant amount of effort has been made to distinguish sequencing error in environmental samples from true rare variants; much of this work has played out in the OTU inference literature \citep{quince2009accurate,quince2011removing,bragg2012fast,edgar2013uparse} as it such errors are especially problematic there.
With the exception of this last reference, OTU inference is not considered to be a phylogenetic problem but rather something to be performed before phylogenetic inference begins.

@@

As described above, \citet{morgan2010metagenomic} showed that various microbes have different DNA extraction efficiencies, meaning that the representation of marker gene sequences is not representative of the actual communities.
Furthermore, there was no clear taxonomic signal in their observations of the variability of extraction efficiency, which seems to preclude a correction strategy based on ``species-tree'' phylogenetic modeling.
However, presumably \textit{something} about their genome is determining extraction efficiency; it would be interesting and useful to search for the genetic determinants.
As described above, abundances are commonly used as part of community comparison, thus a better quantification of error in those observations of abundance would be a great help.

Noah-- AFAIK nobody has done phylogenetic modeling of primer bias, which would seem to be very useful! Couldn't we check for phylogenetic signal and if there is some, do modeling? I mean, there has got to be some phylogenetic signal, right?

@@

Opportunities for clinical applications may present themselves, but only some of these will be sequencing based.
For example, routine 16S sequencing is likely soon to be completely replaced by Matrix-Assisted Laser Desorption Ionization--Time of Flight (MALDI-TOF) mass spectrometry for identification of a single known microbe grown in culture \citep{clark2013matrix}.
However, for diagnoses that are on the level of microbial communities, sequencing and consequent analysis methods will still be required \citep[see review in][]{Rogers2013271}.
Inexpensive whole-genome sequencing will certainly have a profound impact on clinical practice and epidemiological studies \citep{didelot2012transforming}.
For all of these measures, it will be important to have rigorous means of quantifying uncertainty for robust diagnostic applications.
